Global sensitivity analysis of model uncertainty in aeroelastic wind turbine models

### 2020-03-03 Papers/Torque_Paper/article_IOP.tex
final Torque Paper as submitted

\title{A global sensitivity analysis of model uncertainty in aeroelastic wind turbine models}

@@

\begin{abstract}
A framework is presented for performing global sensitivity analysis of model parameters associated with the Blade Element Momentum (BEM) models. The Sobol indices based on a sparse polynomial  expansion are used as a measure of global sensitivities. The sensitivity analysis workflow is developed using the uncertainty quantification toolbox \texttt{UQLab} that is integrated with the ECN's \texttt{Aero-Module} aeroelastic code. Sensitivity studies are performed on the NM80 wind turbine model from the DANAERO project.
\end{abstract}

@@

%\section*{Acknowledgements}

>>>>

\title{Global sensitivity analysis of model uncertainty in aeroelastic wind turbine models}

@@

\begin{abstract}
A framework is presented for performing global sensitivity analysis of model parameters associated with the Blade Element Momentum (BEM) models.  Sobol indices based on adaptive sparse polynomial expansions are used as a measure of global sensitivities. The sensitivity analysis workflow is developed using the uncertainty quantification toolbox \texttt{UQLab} that is integrated with TNO's \texttt{Aero-Module} aeroelastic code. Uncertainties in chord, twist, and lift- and drag-coefficients have been parametrized through the use of NURBS curves. Sensitivity studies are performed on the NM80 wind turbine model from the DanAero project, for a case with 19 uncertainties in both model and geometry. The combination of parametrization and sparse adaptive polynomial chaos yields a new efficient framework for global sensitivity analysis of aeroelastic wind turbine models, paving the way to effective model calibration. 
\end{abstract}

@@

\section*{Acknowledgements}
The research was sponsored by the Topsector Energy Subsidy from the Dutch Ministry of Economic Affairs and Climate.

### 2020-03-03 Papers/Torque_Paper/content.tex
final Torque Paper as submitted

\section{Introduction}
Aeroelastic models such as the Blade Element Momentum (BEM) models \cite{HandBook} play a critical role in the design, development, and optimization of modern wind turbines. A large number of BEM models have been developed to predict turbine responses such as the structural loads and power output \cite{Vorpahl2013}.

@@

The goal of this work is to perform a systematic assessment of the uncertainty in model parameters in BEM models. We approach this by performing a global sensitivity analysis based on the Sobol expansion approach, which decomposes the total variance of the quantity of interest (model output) into contributions from individual parameters and their combinations, similar to \cite{Echeverria2017,Murcia2018, Rinker2016a}. We will employ the uncertainty quantification toolbox \texttt{UQLab} \cite{uqlab}, which computes the Sobol indices based on a sparse polynomial chaos expansion. The main novelty of the work lies in creating a generic framework for studying parametric uncertainteis, including both model uncertainties and geometric uncertainties. This framework for sensitivity analysis is part of the so-called \textit{WindTrue} project, in which the long-term goal is the development of calibrated BEM models. Anticipating on this goal, we are using the NM80 wind turbine model from the Tjaereborg wind farm, for which extensive measurement data is available through the DanAero project \cite{Troldborg2013} and IEA Task 29. The sensitivity analysis will make clear which of the model parameters are to be used in the calibration phase.

@@

\section{Aeroelastic model uncertainties}\label{sec:model_description}
%\subsection{BEM model uncertainties}
State-of-the-art wind-turbine models simulate the aeroelastic behavior of wind turbines by combining the concept of momentum conservation of the flow (BEM theory) with the equations of motion for the structure, possibly extended with the hydrodynamics of the sea and control algorithms \cite{Vorpahl2013}. In this work, we concentrate on the first aspect, namely the prediction of flow and blade forces as given by the BEM method. The particular BEM code that we use is ECN's \texttt{Aero-Module} \cite{Boorsma2012}. For the case of a rigid turbine, with a uniform inflow field, the main uncertainties that arise in the BEM formulation are \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurements or 2D codes like \texttt{XFOIL};
%\item Sheared inflow
\item The assumption of locally 2D flow along the blade is also invalid at the tip; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modeled via \textit{a dynamic stall model} and associated time constant;
\item Yawed inflow is not naturally included and typically incorporated via \textit{a wake correction}.
%\item Dynamic wake - Oye time lag constant
%\item Tower shadow constants " Potential flow / Downwind Empirical model
\end{itemize}
In this paper we will focus on the first uncertainty -- uncertainty in lift and drag coefficients -- but we stress that the framework and methodology are directly applicable to other BEM model uncertainties as well. In a parallel paper, see \cite{Caboni2020}, the effect of other uncertainties such as the dynamic stall model will be investigated. Next to these model uncertainties we will consider geometric uncertainties: the chord and twist distribution along the blade.

@@

\subsection{Perturbing the NURBS curves}
Given the reference values of $\mathbf{c}$, it becomes straightforward to generate perturbed curves: we sample a uniform random variable to perturb the baseline control points and use these to compute a perturbed curve $\tilde{S}(r)$:
\begin{equation}\label{eqn:perturb_c}
\tilde{S} (r) = \sum_{i=1}^{n} \tilde{c}_{i} B_{i,p}(r) \qquad \text{where} \qquad \tilde{c}_{i} = c_{i} (1 + \Delta_{i}),
\end{equation}
and the perturbation is randomly sampled as $\Delta_{i} \sim \mathcal{U}[-0.05,0.05]$. This equation is applicable to both the twist- and chord distribution, where $\tilde{c}_{i}$ can denote either the value of the twist control points, $\text{Tw}_i$, or of the chord control points, $\text{Ch}_{i}$. For the lift- and drag polars, we use the following parametrization:
\begin{equation}\label{eqn:perturb_cl}
\tilde{S} (r,\alpha) = \sum_{i=1}^{n} \tilde{c}_{i}(\alpha) B_{i,p}(r) \qquad \text{where} \qquad \tilde{c}_{i}(\alpha) = c_{\text{ref},i} (\alpha) (1 + \Delta_{i}),
\end{equation}
where $c_{i}(\alpha)$ can denote either the lift coefficient $\text{Cl}_{i}$ or drag coefficient $\text{Cd}_{i}$, and the subscript $\text{ref}$ denotes the reference curve. % $c_{\text{ref},i}(\alpha)$ can denote either the reference lift coefficient $C_{l,\text{ref},i}(\alpha)$ or drag coefficient $C_{d,\text{ref},i}(\alpha)$ at section $i$.

@@

\subsection{Sobol analysis}
We employ the variance-based Sobol decomposition to perform global sensitivity analysis. This approach allows for quantification of the relative importance of the input parameters on a scale of $[0,1]$ known as \emph{Sobol indices}. In a Sobol analysis, we express the total variance of the output in term of contributions from individual parameters and their combinations. Consider a generic BEM model $f(\mathbf{z})$, depending on uncertain parameters $\mathbf{z}$ and returning a quantity of interest $Y$ (e.g.\ power, loads, moments):
\begin{equation}\label{nonlinear_model}
Y = f({\mathbf{z}}),
\end{equation}
where ${\mathbf{z}} = [z_1, z_2, ..., z_M]\in \mathcal{D}_{\mathbf{z}}\in \mathbb{R}^M$ is a vector of random variables (parameters). For ease of notation, we assume these input parameters to be uniformly distributed i.e. $z_i \sim \mathcal{U}(0,1)$ and the support of the input set is $\mathcal{D}_{\mathbf{z}} = [0,1]^M$ where $M$ is the total number of input parameters. The Sobol decomposition is based on the following hierarchical representation \cite{RSmith}:
\begin{equation}\label{sobol_decomp}
f(z_1, z_2, ..., z_M) = f_0+\sum_{i=1}^M f_i(z_i) + \sum_{1\leq i<j\leq M} f_{ij}(z_i,z_j) + ... + f_{1, 2, ..., M}(z_1, z_2, ..., z_M).
\end{equation}
Here the zeroth-order function $f_0$ is the mean response of $f$, the first-order univariate functions $f_i(z_i)$ quantify independent contributions due to each random input, and second-order functions $f_{i,j}(z_i,z_j)$ represent the effect of interaction between $z_i$ and $z_j$ on the response $f$. Formally, the zeroth-, first- and second-order terms are defined as:
\begin{align}\label{sobol_terms}
f_0 = \int_{\mathcal{D}_{\mathbf{z}}}f(\mathbf{z})d\mathbf{z}, \qquad  f_i(z_i)= \int_{{\mathcal{D}_{\mathbf{z}}}^{M-1}}f(\mathbf{z})d\mathbf{z}_{\sim \{i\}} - f_0,\\
f_{ij}(z_i,z_j) = \int_{{\mathcal{D}_{\mathbf{z}}}^{M-2}}f(\mathbf{z})d\mathbf{z}_{\sim \{i,j\}} - f_i(z_i) - f_j(z_j) - f_0,
\end{align}
where ${\mathcal{D}_{\mathbf{z}}}^{M-1} = [0,1]^{M-1}$ and the notation $\mathbf{z}_{\sim \{i,j\}}$ denotes the vector having all the components of $\mathbf{z}$ except in the set $\{i,j\}$. Higher-order terms are interpreted in a similar manner. Note that \eqref{sobol_decomp} is only valid for independent input parameters. We consider a second-order expansion of \eqref{sobol_decomp}, which is generally sufficient for the purpose of this article.

@@

\subsection{PCE-based Sobol indices computation}
A PCE approximation $f^{K}(\mathbf{z})$ to the computational model $Y=f(\mathbf{z})$ can be defined as a weighted sum of multivariate polynomials in $\mathbf{z}$ \cite{RSmith}:
\begin{equation}\label{eq:PCE}
Y= f(\mathbf{z}) \approx f^{K}(\mathbf{z}) = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}),
\end{equation}
where $\mathbf{k}\in \mathbb{N}_0^M$ is multi-index with magnitude $|\mathbf{k}| = k_1+k_2 + ... + k_M$ and $\Psi_{\mathbf{k}}(\mathbf{z})$ a multivariate polynomial basis, computed using a tensor product of univariate polynomials $\psi_{k_i}^{(i)}(z_i)$
\begin{equation}
\Psi_{\mathbf{k}}(\mathbf{z}) := \prod_{i=1}^M\psi_{k_i}^{(i)}(z_i),
\end{equation}
where $k_i$ is the order of the univariate polynomial. The choice of univariate orthogonal polynomial $\psi_{k_i}$ depends on the type of the random variable $z_i$. For example, for uniform random variables the Legendre family of polynomials is used. 
%The multivariate polynomials $\Psi_{\mathbf{k}}(\mathbf{z})$ is orthogonal with respect to the joint probability distribution of random vector $\mathbf{z}$. 
The total number of terms $K$ in \eqref{eq:PCE} depends on the truncation scheme; in a standard truncation scheme  all polynomials in $M$ random variable with order less than or equal to $p$ are used. In this work, a more advanced truncation scheme known as the hyperbolic truncation scheme is used \cite{BlatmanThesis}. 
%\textcolor{red}{In this work, we use the hyperbolic truncation scheme with a $qnorm = 0.75$.}

@@

\subsubsection{Ordinary Least Squares}
%Instead of using a projection-based method one can also use least squares based approaches to compute the coefficients \cite{Berveiller2006}. 
The PCE expansion \eqref{eq:PCE} can be expressed in terms of the following vector-matrix product
\begin{equation}\label{eq:LSM}
Y = f^{K}(\mathbf{z}) +\varepsilon_K = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}) + \varepsilon_K = \mathbf{w}^T\boldsymbol{\Psi}(\mathbf{z}) +\varepsilon_K,
\end{equation} 
where $\mathbf{w} = \{w_0,w_1,...,w_K\}^T$ is the coefficient vector, $\boldsymbol{\Psi}(\mathbf{z}) = \{\Psi_0(\mathbf{z}), \Psi_1(\mathbf{z}), ..., \Psi_K(\mathbf{z})\}^T$ is the matrix containing multivariate polynomials in $\mathbf{z}$, and $\varepsilon_K$ is the approximation error. The least-squares minimization problem can then be formulated as:
\begin{equation}\label{eq:PCE_OLS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right].
 \end{equation}
The coefficient vector $\mathbf{\overline{w}}$ can be computed using the Ordinary Least Squares (OLS) method. Given $N$ model evaluations $\mathbf{Y} = \{f(\mathbf{z}_1), f(\mathbf{z}_2), ..., f(\mathbf{z}_N)\}$, we can obtain the coefficients as:
\begin{equation}\label{eq:OLS}
\overline{\mathbf{w}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y},
\end{equation}
where $\mathbf{X}\in \mathbb{R}^{N\times K}$ is the design matrix where $\mathbf{X}_{i,j} = \Psi_{j}(\mathbf{z}_i)$, $i = 1,2, ..., N$ and $j = 0,1,2,..., K$. Although the OLS method allows to compute the coefficients using a limited number of model evaluations, it may result in a PCE model that consists of high-order interaction terms that may lead to overfitting. For many engineering problems, low-order interactions between the input variables are most important. Therefore, the least squares form in \eqref{eq:PCE_OLS} can be regularized such that the minimization results in a low-order sparse PCE model, as discussed next.

\subsubsection{Least Angle Regression}
The LARS algorithm \cite{efron2004} is one of the methods that can be applied to compute sparse PCE models \cite{BlatmanThesis}. The key idea is to find the best set of polynomials from a given candidate set that minimizes the regularized form of \eqref{eq:PCE_OLS}:
\begin{equation}\label{regularized_LS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right] + \lambda ||\mathbf{w}||_1.
\end{equation}
The LARS algorithm in the context of PCE can be summarized as follows. We start with all PCE coefficients for the candidate polynomials set to zero, and find the polynomial, say $\Psi_1(\mathbf{w})$ most correlated with the model evaluations. This polynomial enters the active set. Next, we take the largest step in the direction of $\Psi_1(\mathbf{w})$ until some other polynomial $\Psi_2(\mathbf{w})$ from the candidate set, has as much correlation with the current residual. The second polynomial is now also added to the active set. In the next step, we move both the coefficients in the active set towards their least-square values until a third polynomial from the candidate set shows as much correlation as the current PCE model based on two polynomials shows with the model evaluations. The number of iterations is given by the minimum of $\{K,N\}$. After every iteration an a-posteriori error, for instance the Leave-One-Out (LOO) cross-validation error, is computed. The active set with the smallest LOO is then chosen as the best sparse PCE model.

One of the main advantage of the LARS algorithm is that it works well in cases when $K>>N$, i.e. the number of polynomials in the candidate set is much larger than the number of model evaluations.

@@

%\subsection{NM80}
\subsection{Operating conditions}
As a reference wind turbine, we consider the 2MW NM80 onshore turbine from the DANAERO project \cite{Troldborg2013} with a blade radius of 38.8 m and a hub height of 57.19 m with the mean wind speed at hub height set as 6.1m/s. Furthermore, the RPM is set at 12.3$/\text{min.}$, the blade pitch angle at 0.15 $\deg$ and the turbine yaw angle is zero.

@@

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm}, clip, scale = 0.22]{convergence_chord1_4.eps}{a}
        \caption{$M=4, Ch1-Ch4$}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.22]{convergence_chord1_6.eps}
        \caption{$M=6, Ch1-Ch6$}
    \end{subfigure}
      ~ 
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.22]{convergence_chord1_8.eps}
        \caption{ $M=8, Ch1-Ch8$}
 \end{subfigure}
    \caption{Convergence of OLS and LARS with increasing number of uncertain inputs $M$ and model evaluations $N$.}
\label{convergence_dim}
\end{figure*}

@@

\subsection{Sensitivity analysis results}
In Fig.~\ref{fig:sensitivity_analysis_power} and \ref{fig:sensitivity_analysis_force}, we show the total order Sobol indices as a measure of the sensitivity of the different geometric and model parameters on the power output and the axial force, respectively, computed using LARS with $M=19$ uncertain parameters chosen from Fig.~\ref{perturbed_samples}. We found that $N=1024$ model evaluations were sufficient to achieve converged indices. In this case, we observe that the uncertainty in model parameters is at least as important as the uncertainty in geometric parameters. Furthermore, within the geometric parameters, the chord distribution shows a significantly higher sensitivity than the twist variables. Lastly, we observe that the axial force is basically only sensitive to the lift coefficient at the outward blade sections, whereas the power is sensitive to both lift and drag coefficients, also at the inward section of the blade.

@@

\FloatBarrier

\section{Conclusions}\label{sec:conclusions}
In this paper we have developed a framework for global sensitivity analysis using Sobol indices computed using sparse adaptive polynomial chaos expansions. The framework can be used for high-dimensional global sensitivity analysis of both geometric and model uncertainties in BEM models. As such, the proposed framework is generic and can deal with many types of (parametric) uncertainties, of which a few were investigated in this paper: twist, chord, lift coefficient and drag coefficient. By using NURBS parameterizations of chord, twist, lift and drag as a function of the blade radius, the number of uncertainties could be limited to 19. The results highlight amongst others the importance of the lift coefficient, especially for the axial force prediction. The identified sensitivities will be utilized in future work to develop calibrated BEM models with built-in uncertainty estimates, as part of the \textit{WindTrue} project, and used to better understand the differences between BEM models and measurement data as studied in IEA Task 29.

>>>>

\section{Introduction}
Aeroelastic models such as the Blade Element Momentum (BEM) models \cite{HandBook} play a critical role in the design, development, and optimization of modern wind turbines. A large number of BEM models have been developed to predict turbine responses such as structural loads and power output \cite{Vorpahl2013}.

@@

The goal of this work is to perform a systematic assessment of the uncertainty in model parameters in BEM models. We approach this by performing a global sensitivity analysis based on the Sobol expansion approach, which decomposes the total variance of the quantity of interest (model output) into contributions from individual parameters and their combinations, similar to \cite{Echeverria2017,Murcia2018, Rinker2016a}. We will employ the uncertainty quantification toolbox \texttt{UQLab} \cite{uqlab}, which computes the Sobol indices based on a sparse polynomial chaos expansion. The main novelty of the work lies in creating a generic framework that enables an advanced (global) sensitivity analysis of a relatively high number of uncertain parameters (e.g.\ 10-20), including both model and geometric uncertainties. The two key ingredients for this are efficient parameterization and the adaptive sparse polynomial expansion. This framework for sensitivity analysis is part of the so-called \textit{WindTrue} project, in which the long-term goal is the development of calibrated BEM models. Anticipating on this goal, we are using the NM80 wind turbine model from the Tjaereborg wind farm, for which extensive measurement data is available through the DanAero project \cite{Troldborg2013} and IEA Task 29  Phase IV (Phase IV is a follow-up of Task 29 Phase III \cite{Boorsma2018} that is built around the DanAero experiment). The sensitivity analysis will make clear which of the model parameters are to be used in the calibration phase.

@@

\section{Aeroelastic model uncertainties}\label{sec:model_description}
%\subsection{BEM model uncertainties}
State-of-the-art wind-turbine models simulate the aeroelastic behavior of wind turbines by combining the concept of momentum conservation of the flow (BEM theory) with the equations of motion for the structure, possibly extended with the hydrodynamics of the sea and control algorithms \cite{Vorpahl2013}. In this work, we concentrate on the first aspect, namely the prediction of flow and blade forces as given by the BEM method. The particular BEM code that we use is TNO's \texttt{Aero-Module} \cite{Boorsma2012}. For the case of a rigid turbine, with a uniform inflow field, the main uncertainties that arise in the BEM formulation are \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurements or 2D codes like \texttt{XFOIL};
%\item Sheared inflow
\item The assumption of an azimuthally uniform induction distribution; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modelled via \textit{a 3D correction or rotational augmentation model};
\item Yawed inflow is not naturally included and incorporated via \textit{a skewed wake correction}.
%\item Dynamic wake - Oye time lag constant
%\item Tower shadow constants " Potential flow / Downwind Empirical model
\end{itemize}
In this paper we will focus on the first uncertainty -- uncertainty in lift and drag coefficients -- but we stress that the framework and methodology are directly applicable to other BEM model uncertainties as well. In a parallel paper, see \cite{Caboni2020}, the effect of other uncertainties such as the unsteady airfoil aerodynamics model will be investigated. Next to these model uncertainties we will consider geometric uncertainties: the chord and twist distribution along the blade, to show the generic applicability of the framework.

@@

\subsection{Perturbing the NURBS curves}
Given the reference values of $\mathbf{c}$, it becomes straightforward to generate perturbed curves: we sample a uniform random variable to perturb the baseline control points and use these to compute a perturbed curve $\tilde{S}(r)$:
\begin{equation}\label{eqn:perturb_c}
\tilde{S} (r) = \sum_{i=1}^{n} \tilde{c}_{i} B_{i,p}(r) \qquad \text{where} \qquad \tilde{c}_{i} = c_{i} (1 + \Delta_{i}),
\end{equation}
and the perturbation is uniformly sampled as $\Delta_{i} \sim \mathcal{U}[-0.05,0.05]$. Other distributions can be specified in the framework; in future work we will further investigate how to choose these, for example based on design standards. This equation is applicable to both twist- and chord distributions: $\tilde{c}_{i}$ can denote either the value of the twist control points, $\text{Tw}_i$, or of the chord control points, $\text{Ch}_{i}$. For the lift- and drag polars, we use the following parametrization:
\begin{equation}\label{eqn:perturb_cl}
\tilde{S} (r,\alpha) = \sum_{i=1}^{n} \tilde{c}_{i}(\alpha) B_{i,p}(r) \qquad \text{where} \qquad \tilde{c}_{i}(\alpha) = c_{\text{ref},i} (\alpha) (1 + \Delta_{i}),
\end{equation}
where $c_{i}(\alpha)$ can denote either the lift coefficient $\text{Cl}_{i}$ or drag coefficient $\text{Cd}_{i}$.%, and the subscript $\text{ref}$ denotes the reference curve. % $c_{\text{ref},i}(\alpha)$ can denote either the reference lift coefficient $C_{l,\text{ref},i}(\alpha)$ or drag coefficient $C_{d,\text{ref},i}(\alpha)$ at section $i$.

@@

\subsection{Sobol analysis}
We employ the variance-based Sobol decomposition to perform global sensitivity analysis. This approach allows for quantification of the relative importance of the input parameters known as \emph{Sobol indices}. In a Sobol analysis, we express the total variance of the output in term of contributions from individual parameters and their combinations. Consider a generic BEM model $f(\mathbf{z})$, depending on uncertain parameters $\mathbf{z}$ and returning a quantity of interest $Y$ (e.g.\ power, loads, moments):
\begin{equation}\label{nonlinear_model}
Y = f({\mathbf{z}}),
\end{equation}
where ${\mathbf{z}} = [z_1, z_2, ..., z_M]\in \mathcal{D}_{\mathbf{z}}\in \mathbb{R}^M$ is a vector of random variables (parameters). For ease of notation, we assume these input parameters to be uniformly distributed i.e. $z_i \sim \mathcal{U}(0,1)$ and the support of the input set is $\mathcal{D}_{\mathbf{z}} = [0,1]^M$ where $M$ is the total number of input parameters. The Sobol decomposition is based on the following hierarchical representation \cite{RSmith}:
\begin{equation}\label{sobol_decomp}
f(z_1, z_2, ..., z_M) = f_0+\sum_{i=1}^M f_i(z_i) + \sum_{1\leq i<j\leq M} f_{ij}(z_i,z_j) + ... + f_{1, 2, ..., M}(z_1, z_2, ..., z_M).
\end{equation}
Here the zeroth-order function $f_0$ is the mean response of $f$, the first-order univariate functions $f_i(z_i)$ quantify independent contributions due to each random input, and second-order functions $f_{i,j}(z_i,z_j)$ represent the effect of interaction between $z_i$ and $z_j$ on the response $f$. Formally, the zeroth-, first- and second-order terms are defined as:
\begin{align}\label{sobol_terms}
f_0 = \int_{\mathcal{D}_{\mathbf{z}}}f(\mathbf{z})d\mathbf{z}, \qquad  f_i(z_i)= \int_{{\mathcal{D}_{\mathbf{z}}}^{M-1}}f(\mathbf{z})d\mathbf{z}_{\sim \{i\}} - f_0,\\
f_{ij}(z_i,z_j) = \int_{{\mathcal{D}_{\mathbf{z}}}^{M-2}}f(\mathbf{z})d\mathbf{z}_{\sim \{i,j\}} - f_i(z_i) - f_j(z_j) - f_0,
\end{align}
where ${\mathcal{D}_{\mathbf{z}}}^{M-1} = [0,1]^{M-1}$ and the notation $\mathbf{z}_{\sim \{i,j\}}$ denotes the vector having all the components of $\mathbf{z}$ except in the set $\{i,j\}$. Higher-order terms are interpreted in a similar manner. Note that \eqref{sobol_decomp} is only valid for independent input parameters. We consider a second-order expansion of \eqref{sobol_decomp}, which is generally sufficient for the purpose of this article.

@@

\subsection{PCE-based Sobol indices computation}
A PCE approximation $f^{K}(\mathbf{z})$ to the computational model $Y=f(\mathbf{z})$ can be defined as a weighted sum of multivariate polynomials in $\mathbf{z}$ \cite{RSmith}:
\begin{equation}\label{eq:PCE}
Y= f(\mathbf{z}) \approx f^{K}(\mathbf{z}) = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}),
\end{equation}
where $\mathbf{k}\in \mathbb{N}_0^M$ is multi-index with magnitude $|\mathbf{k}| = k_1+k_2 + ... + k_M$ and $\Psi_{\mathbf{k}}(\mathbf{z})$ a multivariate polynomial basis, computed using a tensor product of univariate polynomials $\psi_{k_i}^{(i)}(z_i)$, 
%\begin{equation}
%\Psi_{\mathbf{k}}(\mathbf{z}) := \prod_{i=1}^M\psi_{k_i}^{(i)}(z_i),
%\end{equation}
where $k_i$ is the order of the univariate polynomial. The choice of univariate orthogonal polynomial $\psi_{k_i}$ depends on the type of the random variable $z_i$. For example, for uniform random variables the Legendre family of polynomials is used. 
%The multivariate polynomials $\Psi_{\mathbf{k}}(\mathbf{z})$ is orthogonal with respect to the joint probability distribution of random vector $\mathbf{z}$. 
The total number of terms $K$ in \eqref{eq:PCE} depends on the truncation scheme; in a standard truncation scheme  all polynomials in $M$ random variable with order less than or equal to $p$ are used. In this work, a more advanced truncation scheme known as the hyperbolic truncation scheme is used \cite{BlatmanThesis}. 
%\textcolor{red}{In this work, we use the hyperbolic truncation scheme with a $qnorm = 0.75$.}

@@

\subsubsection{Ordinary Least Squares}
%Instead of using a projection-based method one can also use least squares based approaches to compute the coefficients \cite{Berveiller2006}. 
The PCE expansion \eqref{eq:PCE} can be expressed in terms of the following vector-matrix product
\begin{equation}\label{eq:LSM}
Y = f^{K}(\mathbf{z}) +\varepsilon_K = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}) + \varepsilon_K = \mathbf{w}^T\boldsymbol{\Psi}(\mathbf{z}) +\varepsilon_K,
\end{equation} 
where $\mathbf{w} = \{w_0,w_1,...,w_K\}^T$ is the coefficient vector, $\boldsymbol{\Psi}(\mathbf{z}) = \{\Psi_0(\mathbf{z}), \Psi_1(\mathbf{z}), ..., \Psi_K(\mathbf{z})\}^T$ is the matrix containing multivariate polynomials in $\mathbf{z}$, and $\varepsilon_K$ is the approximation error. The least-squares minimization problem can then be formulated as:
\begin{equation}\label{eq:PCE_OLS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right].
 \end{equation}
The coefficient vector $\mathbf{\overline{w}}$ can be computed using the Ordinary Least Squares (OLS) method. Given $N$ model evaluations $\mathbf{Y} = \{f(\mathbf{z}_1), f(\mathbf{z}_2), ..., f(\mathbf{z}_N)\}$, we can obtain the coefficients as $\overline{\mathbf{w}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y}$, where $\mathbf{X}\in \mathbb{R}^{N\times K}$ is the design matrix where $\mathbf{X}_{i,j} = \Psi_{j}(\mathbf{z}_i)$, $i = 1,2, ..., N$ and $j = 0,1,2,..., K$. Although the OLS method allows to compute the coefficients using a limited number of model evaluations, it may result in a PCE model that consists of high-order interaction terms that may lead to overfitting. For many engineering problems, low-order interactions between the input variables are most important. Therefore, the least squares form in \eqref{eq:PCE_OLS} can be regularized such that the minimization results in a low-order sparse PCE model, as discussed next.

\subsubsection{Least Angle Regression}
The LARS algorithm \cite{efron2004} is one of the methods that can be applied to compute sparse PCE models \cite{BlatmanThesis}. The key idea is to find the best set of polynomials from a given candidate set that minimizes the regularized form of \eqref{eq:PCE_OLS}:
\begin{equation}\label{regularized_LS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right] + \lambda ||\mathbf{w}||_1.
\end{equation}
The LARS algorithm in the context of PCE can be summarized as follows. We start with all PCE coefficients for the candidate polynomials set to zero, and find the polynomial, say $\Psi_1(\mathbf{w})$ most correlated with the model evaluations. This polynomial enters the active set. Next, we take the largest step in the direction of $\Psi_1(\mathbf{w})$ until another polynomial $\Psi_2(\mathbf{w})$ from the candidate set, has as much correlation with the current residual. The second polynomial is now also added to the active set. In the next step, we move both the coefficients in the active set towards their least-square values until a third polynomial from the candidate set shows as much correlation as the current PCE model based on two polynomials shows with the model evaluations. The number of iterations is given by the minimum of $\{K,N\}$. After every iteration an a-posteriori error, for instance the Leave-One-Out (LOO) cross-validation error, is computed. The active set with the smallest LOO is then chosen as the best sparse PCE model. One of the main advantage of the LARS algorithm is that it works well in cases when $K>>N$, i.e. the number of polynomials in the candidate set is much larger than the number of model evaluations.

@@

%\subsection{NM80}
\subsection{Operating conditions}
As a reference wind turbine, we consider the 2MW NM80 onshore turbine from the DANAERO project \cite{Troldborg2013} with a blade radius of 38.8 m and a hub height of 57.19 m with the mean wind speed at hub height set as 6.1m/s. Furthermore, the rotational speed is set at 12.3 rpm, the blade pitch angle at 0.15 $\deg$ and the turbine yaw angle is zero.

@@

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm}, clip, scale = 0.22]{convergence_chord1_4.eps}{a}
        \caption{$M=4, Ch1-Ch4$}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.22]{convergence_chord1_6.eps}
        \caption{$M=6, Ch1-Ch6$}
    \end{subfigure}
      ~ 
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.22]{convergence_chord1_8.eps}
        \caption{ $M=8, Ch1-Ch8$}
 \end{subfigure}
    \caption{Convergence of OLS and LARS with increasing number of uncertain inputs $M$ (chord control points) and model evaluations $N$.}
\label{convergence_dim}
\end{figure*}

@@

\subsection{Sensitivity analysis results}
In Fig.~\ref{fig:sensitivity_analysis_power} and \ref{fig:sensitivity_analysis_force}, we show the total order Sobol indices as a measure of the sensitivity of the different geometric and model parameters on the power output and the axial force, respectively, computed using LARS with $M=19$ uncertain parameters chosen from Fig.~\ref{perturbed_samples}. We found that $N=1024$ model evaluations were sufficient to achieve converged indices. In this case, we observe that the uncertainty in model parameters is at least as important as the uncertainty in geometric parameters. Furthermore, within the geometric parameters, the chord distribution shows a significantly higher sensitivity than the twist variables. Lastly, we observe that the axial force is basically only sensitive to the lift coefficient at the outward blade sections, whereas the power is sensitive to both lift and drag coefficients, also at the inward section of the blade. We note that these findings are testcase-specific and might not generalize directly to other turbines or other operational settings. Furthermore, the reported sensitivities depend on the variance that is chosen for the uncertain parameters; in future work we will relate these to values known from design standards.

@@

\section{Conclusions}\label{sec:conclusions}
In this paper we have developed a framework for global sensitivity analysis using Sobol indices computed using adaptive sparse polynomial chaos expansions. The framework can be used for high-dimensional global sensitivity analysis of both geometric and model uncertainties in BEM models. As such, the proposed framework is generic and can deal with many types of (parametric) uncertainties, of which a few were investigated in this paper: twist, chord, lift coefficient and drag coefficient. By using NURBS parameterizations of chord, twist, lift and drag as a function of the blade radius, the number of uncertainties was limited but remains considerable (19 uncertain variables). Our approach (adaptive sparse polynomial chaos expansion) is able to efficiently quantify this high-dimensional problem. The results highlight amongst others the importance of the lift coefficient, especially for the axial force prediction. The identified sensitivities will be utilized in future work to develop calibrated BEM models with built-in uncertainty estimates, as part of the \textit{WindTrue} project, and used to better understand the differences between BEM models and measurement data as studied in IEA Task 29.
